Acessibilidade nas fases de Engenharia de Requisitos, Projeto e Codificação de Software: Uma ferramenta de apoio

### 2014-01-09 paper.tex
.

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Rodrigo Gonçalves de Branco%
%\thanks{Please note that the LNCS Editorial assumes that all authors have used
%the western naming convention, with given names preceding surnames. This determines
%the structure of the names in the running heads and the author index.}%
\and Débora Maria Barroso Paiva\and Maria Istela Cagnin}

@@

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Faculdade de Computação - FACOM, \\
Universidade Federal de Mato Grosso do Sul - UFMS,\\
Cidade Universitária, Campo Grande - MS, Brasil\\
\mailsa\\
\url{http://facom.ufms.br/}}

@@

\begin{abstract}
The abstract should summarize the contents of the paper and should
contain at least 70 and at most 150 words. It should be written using the
\emph{abstract} environment.
\keywords{We would like to encourage you to list your keywords within
the abstract section}
\end{abstract}

\section{Introduction}

You are strongly encouraged to use \LaTeXe{} for the
preparation of your camera-ready manuscript together with the
corresponding Springer class file \verb+llncs.cls+. Only if you use
\LaTeXe{} can hyperlinks be generated in the online version
of your manuscript.

The \LaTeX{} source of this instruction file for \LaTeX{} users may be
used as a template. This is
located in the ``authors'' subdirectory in
\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
entitled \texttt{typeinst.tex}. There is a separate package for Word 
users. Kindly send the final and checked source
and PDF files of your paper to the Contact Volume Editor. This is
usually one of the organizers of the conference. You should make sure
that the \LaTeX{} and the PDF files are identical and correct and that
only one version of your paper is sent. It is not possible to update
files at a later stage. Please note that we do not need the printed
paper.

We would like to draw your attention to the fact that it is not possible
to modify a paper in any way, once it has been published. This applies
to both the printed book and the online version of the publication.
Every detail, including the order of the names of the authors, should
be checked before the paper is sent to the Volume Editors.

\subsection{Checking the PDF File}

Kindly assure that the Contact Volume Editor is given the name and email
address of the contact author for your paper. The Contact Volume Editor
uses these details to compile a list for our production department at
SPS in India. Once the files have been worked upon, SPS sends a copy of
the final pdf of each paper to its contact author. The contact author is
asked to check through the final pdf to make sure that no errors have
crept in during the transfer or preparation of the files. This should
not be seen as an opportunity to update or copyedit the papers, which is
not possible due to time constraints. Only errors introduced during the
preparation of the files will be corrected.

This round of checking takes place about two weeks after the files have
been sent to the Editorial by the Contact Volume Editor, i.e., roughly
seven weeks before the start of the conference for conference
proceedings, or seven weeks before the volume leaves the printer's, for
post-proceedings. If SPS does not receive a reply from a particular
contact author, within the timeframe given, then it is presumed that the
author has found no errors in the paper. The tight publication schedule
of LNCS does not allow SPS to send reminders or search for alternative
email addresses on the Internet.

In some cases, it is the Contact Volume Editor that checks all the final
pdfs. In such cases, the authors are not involved in the checking phase.

\subsection{Additional Information Required by the Volume Editor}

If you have more than one surname, please make sure that the Volume Editor
knows how you are to be listed in the author index.

\subsection{Copyright Forms}

The copyright form may be downloaded from the ``For Authors"
(Information for LNCS Authors) section of the LNCS Website:
\texttt{www.springer.com/lncs}. Please send your signed copyright form
to the Contact Volume Editor, either as a scanned pdf or by fax or by
courier. One author may sign on behalf of all of the other authors of a
particular paper. Digital signatures are acceptable.

@@

>>>>

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Rodrigo Gonçalves de Branco%
%\thanks{Please note that the LNCS Editorial assumes that all authors have used
%the western naming convention, with given names preceding surnames. This determines
%the structure of the names in the running heads and the author index.}%
\and Débora Maria Barroso Paiva \and Maria Istela Cagnin}

@@

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Faculdade de Computação - FACOM\\
Universidade Federal de Mato Grosso do Sul - UFMS\\
Cidade Universitária, Campo Grande - MS, Brasil\\
\mailsa\\
\url{http://facom.ufms.br/}}

@@

\begin{abstract}
The abstract should summarize the contents of the paper and should
contain at least 70 and at most 150 words. It should be written using the
\emph{abstract} environment.
\keywords{Accessibility, Requirement Traceability, Software Development
Process, CASE Tool}
%\keywords{We would like to encourage you to list your keywords within
%the abstract section}
\end{abstract}

\section{Introdução}

A Internet se consolidou como um dos principais meios de comunicação, disseminação da informação e fornecimento de serviços nos dias atuais, sendo que determinados tipos de informações são disponibilizados apenas através deste meio. Fornecer o acesso a estas informações para uma imensa variedade de dispositivos, com tecnologias e especificações diferentes, se tornou um desafio. E, principalmente, pessoas com necessidades especiais, devido a suas características inerentes, necessitam de produtos especialmente desenvolvidos para este fim.

Construir produtos acessíveis não é uma tarefa fácil e este assunto é alvo de muitas pesquisas \cite{lazar:04,brajnik:06,zeng:05}, existindo propostas específicas para integrar usabilidade e acessibilidade aos processos de Engenharia de Software \cite{springerlink:10.1007/978-3-642-02713-0,maia:10}. Contudo, devido a fatores como a recente exposição do tema nos meios de comunicação e a deficiência no treinamento e formação  dos desenvolvedores, muitos deles sequer sabem como codificar para tornar seus produtos acessíveis \cite{kavcic:05,alves:11}.

A utilização de ferramentas que apóiam os profissionais na construção de boas soluções acessíveis é muito comum, pois estas aumentam a produtividade e diminuem o esforço necessário. Estas ferramentas podem variar entre ferramentas CASE, IDEs, frameworks, simuladores, validadores, avaliadores e até mesmo uma mistura destas citadas. Contudo, pesquisas mostram que as ferramentas disponíveis deixam a desejar no quesito de apoio à acessibilidade \cite{trewin:10}.

A situação é agravada pela constatação de que, apesar de existirem vários estudos sobre rastreabilidade de requisitos genéricos \cite{ali:11,gotel:94,soonsongtanee:2010,mader:12}, poucos estudos têm indicado como ocorre a evolução dos requisitos de acessibilidade durante o processo de desenvolvimento das aplicações web \cite{dias:10}, e como os desenvolvedores utilizarão essa informação para construir o produto.

Este artigo apresenta o AccTrace, uma ferramenta CASE desenvolvida como um plugin do Eclipse para, através da rastreabilidade dos requisitos de acessibilidade, entregar ao desenvolvedor informações relevantes para a construção do produto.

@@

\bibitem{lazar:04} Lazar, J., Dudley-Sponaugle, A.,  Greenidge, K.-D.: Improving web accessibility:
a study of webmaster perceptions. Computers in Human Behavior, 20(2), 269--288. (2004)

\bibitem{brajnik:06} Brajnik, G.: Web Accessibility Testing: When the Method Is the Culprit. In: Miesenberger, K., Klaus, J., Zagler, W. L., e Karshmer, A. I. (eds.) ICCHP 2006, vol. 4061, pp. 156--163. Springer (2006)

\bibitem{zeng:05} Parmanto, B. e Zeng, X.: Metric for Web accessibility evaluation. JASIST, 56(13), 1394--1404. (2005)

\bibitem{springerlink:10.1007/978-3-642-02713-0} Moreno, L., Martínez, P., Ruiz-Mezcua, B.: Integrating HCI in a Web Accessibility Engineering Approach. In: Stephanidis, C., (ed.) Universal Access in Human-Computer Interaction. Applications and Services, vol. 5616, pp. 745--754. Springer (2009)

\bibitem{maia:10} Maia, L. S.: Um processo para o desenvolvimento de aplicações Web Acessíveis.
Master Thesis, UFMS (2010)

\bibitem{kavcic:05} Kavcic, A.: Software Accessibility: Recommendations and Guidelines. In: Computer as a Tool. EUROCON, The International Conference, vol. 2, pp 1024--1027. (2005)

\bibitem{alves:11} Alves, D. D.: Acessibilidade no Desenvolvimento de Software Livre. Master Thesis, UFMS (2011)

\bibitem{trewin:10} Trewin, S., Cragun, B., Swart, C., Brezin, J., Richards, J.: Accessibility challenges
and tool features: an IBM Web developer perspective. In: Proceedings of the
2010 International Cross Disciplinary Conference on Web Accessibility. W4A 2010, pp 32:1--32:10. ACM. (2010)

\bibitem{dias:10} Dias, A. L., de Mattos Fortes, R. P., Masiero, P. C., Goularte, R.: Uma Revisão
Sistemática sobre a inserção de Acessibilidade nas fases de desenvolvimento da Engenharia
de Software em sistemas Web. In: Proceedings of the IX Symposium on Human
Factors in Computing Systems. IHC 2010, pp. 39--48. SBC (2010)

\bibitem{ali:11} Ali, N., Gueheneuc, Y., Antoniol, G.: Trust-Based Requirements Traceability.
In Program Comprehension. ICPC 2011, pp. 111--120. IEEE (2011)

\bibitem{gotel:94} Gotel, O. C. Z., Finkelstein, A. C. W.: An analysis of the requirements traceability
problem. In Requirements Engineering, Proceedings of the First International Conference on, pp. 94--101 (1994)

\bibitem{soonsongtanee:2010} Soonsongtanee, S., Limpiyakorn, Y.: Enhancement of requirements traceability
with state diagrams. In: Computer Engineering and Technology (ICCET), 2nd International Conference on, vol. 2, pp. 248--252 (2010)

\bibitem{mader:12} Mader, P., Egyed, A.: Assessing the effect of requirements traceability for software
maintenance. In: Software Maintenance (ICSM), 28th IEEE International Conference on, pp. 171--180 (2012)
